Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2023-05-05 paper/ms.tex
Change to bar locking

%We argue that this can be understood in terms of the halo wake mechanism. 
We interpret this behavior in terms of the halo wake mechanism. In the
\Nbody{} case, a well known phenomenon is that the halo material resonant with
the bar forms a wake, and this wake lags behind \citep{1984MNRAS.209..729T,
1985MNRAS.213..451W, 1992ApJ...400...80H} and exerts a negative torque on the
bar, slowing it down (see Fig.~\ref{fig:wake} below).\footnote{Since the bar is not a solid
body, it is not guaranteed that a negative torque will slow it down - e.g. a
negative torque could shred the bar, reducing its moment of inertia without
changing its pattern speed. However, the bar seems to empirically respond to a
negative torque induced by a halo wake by slowing down.} As the bar slows down,
the location of the resonances in the phase space changes (see Fig.~12 and Table
1 in \citet{2020ApJ...890..117D}) allowing halo material newly resonant with the
bar to participate in the formation of the wake. However, the gas is also a
reliable source of positive torque on the bar, speeding the bar up. In turn,
this stops the location of the resonance from changing such that the halo cannot
reinforce the wake, therefore arresting the process by which the halo can slow
the bar down. We term this process ``halo exhaustion.''

%In the \Nbody{} case, halo material that is
%resonant with the bar will form a wake which lags behind and exerts a negative 
%torque on the bar, slowing it down \citep{1984MNRAS.209..729T,
%1985MNRAS.213..451W, 1992ApJ...400...80H}.\footnote{Since the bar is not a solid
%body, it is not guaranteed that a negative torque will slow it down - e.g. a
%negative torque could shred the bar, reducing its moment of inertia without
%changing its pattern speed. However, the bar seems to empirically respond to a
%negative torque induced by a halo wake by slowing down.} As the bar slows down,
%the location of the resonances in phase space changes, allowing halo material
%newly resonant with the bar to participate in forming the wake. However, the gas
%is a reliable source of positive torque on the bar, speeding the bar up. This
%stops the resonance location from changing such that the halo cannot reinforce
%the wake, arresting the process by which the halo can slow the bar down. We term
%this process ``halo exhaustion.''

This halo exhaustion process is similar to the ``metastability'' effect, which
has been previously discussed in the literature \citep{2003MNRAS.345..406V,
2006ApJ...639..868S}. Finally, we also note that these authors, in particular,
observe that the effects of numerical resolution in the simulations adopted to
explore these mechanisms have yet to be fully explored and could play a role in
the observed phenomenology of the simulations. We plan to address these issues
in future dedicated work.

@@

We argue that the following occurs. First, the bar is not able to slow down
quickly enough due to the positive torque of the infalling gas. This causes
the resonant halo phase space at a particular pattern speed, $\Omega_{\textrm{p},0}$,
to become exhausted and no longer able to support a negative torque. Second,
the gas is still exerting a positive torque on the bar, and therefore the
pattern speed will again increase. Since at higher pattern speeds the halo has
not yet been totally exhausted, the halo will once again be able to exert a
negative torque on the bar. The pattern speed will then settle at a new value
slightly higher than $\Omega_{\textrm{p},0}$ where the gas and halo torques cancel.
Over time, the pattern speed should slowly increase as the halo becomes
progressively exhausted.

@@

We next take the $\tau_{\textrm{gas}}=0$ and $20$ cases from Fig.~\ref{fig:sam}
and plot the halo torque evolution. This result, given in
Fig.~\ref{fig:sam-torque}, is comparable to the lower panel of
Fig.~\ref{fig:prop}. We find that the $\tau_{\textrm{gas}}=0$ case compares
favorably to the \Nbody{} case described in previous sections. The bar exerts a
steady negative torque in this case (blue line). When a gas torque is included
(orange lines), we find that the halo's torque becomes much weaker, similar to
what we found in the \SMUGGLE{} case. The gas applies a steady positive torque
by construction. Therefore, the exhaustion process is reproduced in our simple
semi-analytic model. {\bf Curiously, we do not see the bimodal behavior in
Fig.~\ref{fig:fgas} and described in \citet{2010ApJ...719.1470V}. It is not
presently clear why this is the case.}

>>>>

%We argue that this can be understood in terms of the halo wake mechanism. 
We interpret this behavior in terms of the halo wake mechanism. In the
\Nbody{} case, a well known phenomenon is that the halo material resonant with
the bar forms a wake, and this wake lags behind \citep{1984MNRAS.209..729T,
1985MNRAS.213..451W, 1992ApJ...400...80H} and exerts a negative torque on the
bar, slowing it down (see Fig.~\ref{fig:wake} below).\footnote{Since the bar is not a solid
body, it is not guaranteed that a negative torque will slow it down - e.g. a
negative torque could shred the bar, reducing its moment of inertia without
changing its pattern speed. However, the bar seems to empirically respond to a
negative torque induced by a halo wake by slowing down.} As the bar slows down,
the location of the resonances in the phase space changes (see Fig.~12 and Table
1 in \citet{2020ApJ...890..117D}) allowing halo material newly resonant with the
bar to participate in the formation of the wake. However, the gas is also a
reliable source of positive torque on the bar, speeding the bar up. In turn,
this stops the location of the resonance from changing such that the halo cannot
reinforce the wake, therefore arresting the process by which the halo can slow
the bar down. We term this process ``bar locking.''

This bar locking process is similar to the ``metastability'' effect, which has
been previously discussed in the literature \citep{2003MNRAS.345..406V,
2006ApJ...639..868S}. Finally, we also note that these authors, in particular,
observe that the effects of numerical resolution in the simulations adopted to
explore these mechanisms have yet to be fully explored and could play a role in
the observed phenomenology of the simulations. We plan to address these issues
in future dedicated work.

@@

We argue that the following occurs. First, the bar is not able to slow down
quickly enough due to the positive torque of the infalling gas. This causes the
resonant halo phase space at a particular pattern speed,
$\Omega_{\textrm{p},0}$, to become {\bf mixed} and no longer able to support a
negative torque.\footnote{In the halo wake picture, this corresponds to the wake
becoming fully aligned with the bar.} Second, the gas is still exerting a
positive torque on the bar, and therefore the pattern speed will again increase.
Since at higher pattern speeds the halo has not yet been totally {\bf mixed},
the halo will once again be able to exert a negative torque on the bar. The
pattern speed will then settle at a new value slightly higher than
$\Omega_{\textrm{p},0}$ where the gas and halo torques cancel. Over time, the
pattern speed should slowly increase.

@@

We next take the $\tau_{\textrm{gas}}=0$ and $20$ cases from Fig.~\ref{fig:sam}
and plot the halo torque evolution. This result, given in
Fig.~\ref{fig:sam-torque}, is comparable to the lower panel of
Fig.~\ref{fig:prop}. We find that the $\tau_{\textrm{gas}}=0$ case compares
favorably to the \Nbody{} case described in previous sections. The bar exerts a
steady negative torque in this case (blue line). When a gas torque is included
(orange lines), we find that the halo's torque becomes much weaker, similar to
what we found in the \SMUGGLE{} case. The gas applies a steady positive torque
by construction. Therefore, the locking process is reproduced in our simple
semi-analytic model. {\bf Curiously, we do not see the bimodal behavior in
Fig.~\ref{fig:fgas} and described in \citet{2010ApJ...719.1470V}. It is not
presently clear why this is the case.}
